
```{r}
#| include: true
```

---
title: "Replication code for 'Making, Updating, and Querying Causal Models using CausalQueries'"
date: `r Sys.Date()`
author: Till Tietz, Lily Medina, Georgiy Syunyaev, and Macartan Humphreys
format:
  html: 
    toc: true
    toc-depth: 2
---

Generated by `knitr::spin()` from `code.R`.

Note: final section of code for microbenchmarking is slow
## Set up

```{r}
#| label: preamble
#| include: true

library("tidyverse")
library("CausalQueries")
library("microbenchmark")
library("parallel")
library("future")
library("future.apply")
library("knitr")
library("rstan")

options(kableExtra.latex.load_packages = FALSE)
library("kableExtra")
options(mc.cores = parallel::detectCores())


if (knitr::is_latex_output()) {
  
knitr::knit_hooks$set(source = function(x, options) {


  lines <- unlist(strsplit(x, "\n"))
  prompts <- character(length(lines))
  continuation <- FALSE
  open_brackets <- 0

  for (i in seq_along(lines)) {
    line_trimmed <- trimws(lines[i])

    # Determine if the previous line ends with |> or has unmatched brackets
    if (continuation || open_brackets > 0) {
      prompts[i] <- "+  "
    } else {
      prompts[i] <- "R> "
    }

    # Update brackets count
    open_brackets <- open_brackets +
      stringr::str_count(line_trimmed, "[\\(\\{\\[]") -
      stringr::str_count(line_trimmed, "[\\)\\}\\]]")

    # Check if line ends with pipe operator (|>)
    continuation <- grepl("\\|>\\s*$", line_trimmed)

    # Reset bracket count on empty lines
    if (line_trimmed == "") {
      open_brackets <- 0
      continuation <- FALSE
    }
  }

  # Add exactly two spaces indentation to continuation lines starting with "+"
  formatted_lines <- ifelse(
    prompts == "R> ",
    paste0(prompts, trimws(lines, "left")),
    paste0(prompts, "  ", trimws(lines, "left"))
  )

  paste0("\\begin{CodeInput}\n",
         paste(formatted_lines, collapse = "\n"),
         "\n\\end{CodeInput}\n")
})
}

set.seed(1, "L'Ecuyer-CMRG")
theme_set(theme_bw())

kabble <- function(
    x, 
    align = "c", 
    format = ifelse(knitr::is_html_output(), "html", "latex"), 
    ...) {
  x |>
    knitr::kable(
      format = format,
      digits = 2,
      align = align,
      booktabs = TRUE, 
      linesep = "",
      longtable = TRUE,
      ...) |>
    kableExtra::kable_classic_2(latex_options = c("scale_down"))
}



```
```{r}
#| include: true
```

# Motivating example

```{r}
#| echo: true

data("lipids_data")
lipids_data



```
```{r}
#| include: true
```

# Motivating example

```{r}
#| echo: true
#| eval: true
#| purl: true
#| message: false

lipids_model <-  make_model("Z -> X -> Y; X <-> Y") 



```
```{r}
#| echo: true
#| eval: true
#| purl: true
#| message: false

lipids_model <- update_model(lipids_model, lipids_data)



```
```{r}
#| echo: true
#| eval: true
#| purl: true

lipids_queries <- query_model(lipids_model, queries = list(
  ATE = "Y[X = 1] - Y[X = 0]",
  PoC = "Y[X = 1] - Y[X = 0] :|: X == 0 & Y == 0",
  LATE = "Y[X = 1] - Y[X = 0] :|: X[Z = 1] > X[Z = 0]"),
  using = "posteriors")



```
```{r}
#| label: tbl-lipids
#| tbl-cap: "Replication of \\citet{chickering_clinicians_1996}."
#| echo: true

lipids_queries |>
  dplyr::select(label, query, given, mean, sd, starts_with("cred")) |>
kabble(align = "c")




```
```{r}
#| fig-cap: "Illustration of queries plotted."
#| label: queryplot
#| fig-width: 8
#| fig-height: 3

lipids_queries |> plot()



```
```{r}
#| echo: true
#| include: false
#| message: false

# make_model("Z -> X -> Y; X <-> Y") |> update_model(lipids_data) |>
#   query_model(queries = list(ATE = "Y[X = 1] - Y[X = 0]",
#   PoC  = "Y[X = 1] - Y[X = 0] :|: X == 0 & Y == 0",
#   LATE = "Y[X = 1] - Y[X = 0] :|: X[Z = 1] > X[Z = 0]"),
#   using = "posteriors") |> plot()
# 


```
```{r}
#| include: true
```

# Statistical model

```{r}
#| label: tbl-lipidspar
#| tbl-cap: "Nodal types and parameters for Lipids model."
#| echo: true

with_pars <- 
  lipids_model |>
  set_parameters(param_type = "prior_draw") 

with_pars |> grab("parameters_df") |>
  dplyr::select(node, nodal_type, param_set, param_names, param_value, priors) |> 
  kabble()



```
```{r}
#| include: true
```

# Making models

```{r}
#| echo: true
#| results: markup

model <- make_model("X -> M -> Y <- X")



```
```{r}
#| include: true
```

## Graphing

```{r}
#| label: fig-plots
#| echo: true
#| out-width: 50%
#| fig-height: 6
#| fig-width: 4
#| fig-cap: "Examples of model graphs."
#| fig-subcap:
#|   - "Without options"
#|   - "With options"
#| layout-ncol: 2
#| purl: true

lipids_model |> plot()
lipids_model |> plot(x_coord = 1:3, y_coord = 3:1, textcol = "black",
  textsize = 3, shape = c(15, 16, 16), nodecol = "lightgrey", nodesize = 10)



```
```{r}
#| include: true
```

## Tailoring models

```{r}
#| label: tbl-dof
#| tbl-cap: "Number of different independent parameters (degrees of freedom) for different three-node models."
#| tbl-pos: "b"
#| echo: true
#| eval: true

statements <- list("X -> Y <- W", 
                   "X -> Y <- W; X <-> W", 
                   "X -> Y <- W; X <-> Y; W <-> Y",
                   "X -> Y <- W; X <-> Y; W <-> Y; X <-> W", 
                   "X -> W -> Y <- X",
                   "X -> W -> Y <- X; W <-> Y",
                   "X -> W -> Y <- X; X <-> W; W <-> Y", 
                   "X -> W -> Y <- X; X <-> W; W <-> Y; X <-> Y")

dof <- function(statement) {
  make_model(statement, add_causal_types = FALSE) |>
  grab("parameters_df")  |>
  group_by(param_set) |>
  summarize(n  = n() - 1) |>
  pull(n) |>
  sum()
}



data.frame(
  Model = c(
  "X → Y ← W",
  "X → Y ← W; X ←→ W",
  "X → Y ← W; X ←→ Y; W ←→ Y",
  "X → Y ← W; X ←→ Y; W ←→ Y; X ←→ W",
  "X → W → Y ← X",
  "X → W → Y ← X; W ←→ Y",
  "X → W → Y ← X; X ←→ W; W ←→ Y",
  "X → W → Y ← X; X ←→ W; W ←→ Y; X ←→ Y"
),
  dof = unlist(lapply(statements, dof))
) |> 
  kabble(
    col.names = c("Model", "Degrees of freedom"),
    align = c("l", "c")
  )




```
```{r}
#| echo: true
#| eval: true

model_restricted <- lipids_model |> 
  set_restrictions("X[Z = 1] < X[Z = 0]")



```
```{r}
#| label: set-restrictions1
#| echo: true
#| purl: true

model <- lipids_model |> set_restrictions(
  labels = list(X = "01", Y = c("00", "01", "11")), keep = TRUE)



```
```{r}
#| label: set-restrictions2
#| echo: true
#| purl: true

model <- lipids_model |> set_restrictions(labels = list(Y = "?0"))



```
```{r}
#| label: set-restrictions3
#| echo: true
#| purl: true

model <- lipids_model |> set_restrictions(
  labels = list(Y = c('00', '11')), given = 'X.00')



```
```{r}
#| label: get-priors
#| echo: true
#| eval: true

lipids_model |> inspect("prior_hyperparameters", nodes = "X") 



```
```{r}
#| label: set-priors
#| echo: true
#| purl: true

model <- lipids_model |> set_priors(distribution = "jeffreys")



```
```{r}
#| label: set-priors-custom
#| echo: true
#| eval: true
#| message: false

lipids_model |> set_priors(
  param_names = c("X.10", "X.01"), alphas = 3:4) |> 
  inspect("prior_hyperparameters", nodes = "X")



```
```{r}
#| label: set-priors-statement
#| echo: true
#| eval: true

lipids_model |> set_priors(
  statement = "X[Z = 1] > X[Z = 0]", alphas = 3) |>
  inspect("prior_hyperparameters", nodes = "X")



```
```{r}
#| label: set-priors-flat

query <- make_model("X -> Y") |> 
  set_restrictions(decreasing("X", "Y")) |>
  query_model("Y[X = 1] - Y[X = 0]", using = "priors")



```
```{r}
#| label: get-parameters
#| echo: true
#| eval: true

make_model("X -> Y") |> inspect("parameters")



```
```{r}
#| label: set-parameters
#| echo: true
#| eval: true

make_model("X -> Y") |> set_parameters(
  statement = "Y[X = 1] > Y[X = 0]", parameters = .7) |>
  inspect("parameters")



```
```{r}
#| include: true
```

## Drawing and manipulating data

```{r}
#| label: make-data
#| echo: true
#| eval: true
#| include: true

lipids_model |> make_data(n = 4)



```
```{r}
#| label: make-data-incomplete
#| echo: true
#| eval: true
#| message: false

sample_data <- lipids_model |> make_data(
  n = 8, nodes = list(c("Z", "Y"), "X"), probs = list(1, .5),
  subsets = list(TRUE, "Z == 1 & Y == 0"))



```
```{r}
#| label: collapse-data
#| echo: true
#| eval: true

sample_data |> collapse_data(lipids_model)



```
```{r}
#| include: true
```

# Updating models

```{r}
#| include: true
```

## How the Stan program works

```{r}
#| echo: true
#| eval: true
#| purl: true 
#| include: true

make_model("X -> Y") |> inspect("parameter_mapping") 



```
```{r}
#| include: true
```

## Incomplete and censored data

```{r}
#| echo: true
#| eval: true
#| message: false

data <- data.frame(X = rep(0:1, 5), Y = rep(0:1, 5))
list(uncensored = update_model(make_model("X -> Y"), 
  data = data, refresh = 0),
    censored = update_model(make_model("X -> Y"), data = data, 
    censored_types = c("X1Y0",  "X0Y1"), refresh = 0)) |>
  query_model("Y[X = 1] - Y[X = 0]", using = "posteriors") |>
  dplyr::select(-using)



```
```{r}
#| include: true
```

## Output

```{r}
#| echo: true
#| purl: true
#| message: false

model <- make_model("X -> Y") |> update_model()
posterior <- inspect(model, "posterior_distribution")  



```
```{r}
#| echo: true
#| purl: true
#| message: false

lipids_model <- lipids_model |> update_model(
  keep_fit = TRUE, keep_event_probabilities = TRUE)



```
```{r}
#| eval: true
#| echo: true
#| purl: true
#| message: false

make_model("X -> Y") |> update_model(keep_type_distribution = FALSE, 
  refresh = 0) |> inspect("stan_summary") 



```
```{r}
#| include: true
```

## Convergence problems and diagnostics

```{r}
#| label: preservewwarning
#| warning: false

model <- make_model("X -> M -> Y") |> update_model(
  data = data.frame(X = rep(0:1, 10000), Y = rep(0:1, 10000)), 
  iter = 5000, refresh = 0)



```
```{r}
#| eval: true
#| echo: true
#| purl: true

model



```
```{r}
#| eval: true
#| echo: true
#| purl: true
#| message: false

model <- make_model("X -> Y") |> 
  update_model(refresh = 0, keep_fit = TRUE)



```
```{r}
#| eval: true
#| echo: true
#| purl: true

model |> inspect("stanfit")



```
```{r}
#| include: true
```

# Querying models

```{r}
#| include: true
```

## Calculating factual and counterfactual quantities

```{r}
#| label: realise-outcomes
#| echo: true
#| include: true

make_model("X -> Y") |> realise_outcomes()



```
```{r}
#| label: realise-outcomes-do
#| echo: true
#| include: true

make_model("X -> Y") |> realise_outcomes(dos = list(X = 1))



```
```{r}
#| include: true
```

## Causal syntax

```{r}

make_model("X -> Y")  |> get_query_types("Y == 1")



```
```{r}

make_model("X -> Y") |> get_query_types("Y[X = 0] == 1")



```
```{r}
#| eval: true
#| echo: true

make_model("X1 -> Y <- X2")  |> get_query_types(
  "X1 == 1 & X2 == 1 & (Y[X1 = 1, X2 = 1] > Y[X1 = 0, X2 = 0])")



```
```{r}
make_model("X -> Y") |> get_query_types("Y[X = 1] - Y[X = 0]")



```
```{r}
#| include: true
```

## Quantifying queries

```{r}
#| label: fig-posterior-dist
#| fig-cap: 'Posterior on "Probability $Y$ is increasing in $X$".'
#| fig-pos: "t"
#| fig-align: center
#| out-width: "60%"
#| purl: true
#| message: false

data  <- data.frame(X = rep(0:1, 50), Y = rep(0:1, 50))

model <-  make_model("X -> Y") |> update_model(
  data, iter = 4000, refresh = 0)

model |> grab("posterior_distribution")  |> 
  ggplot(aes(Y.01 - Y.10)) + geom_histogram() 



```
```{r}
#| echo: true

queries <- make_model("X -> Y") |> query_distribution(
  query = list(increasing = "(Y[X = 1] > Y[X = 0])", 
    ATE = "(Y[X = 1] - Y[X = 0])"), using = "priors")



```
```{r}
#| label: fig-case-level-query
#| fig-cap: "Illustration of case-level queries plotted."
#| echo: true
#| eval: true
#| purl: true
#| fig-width: 8
#| fig-height: 3

lipids_model |> query_model(
    query = "Y[X = 1] - Y[X = 0] :|: X == 1 & Y == 1 & Z == 1",
    using = "posteriors") |> plot()



```
```{r}
#| label: fig-new-case-level-query
#| fig-cap: "Illustration of new case-level queries plotted."
#| echo: true
#| eval: true
#| purl: true
#| fig-width: 8
#| fig-height: 3
#| message: false

make_model("X -> M -> Y") |> update_model(
  data.frame(X = rep(0:1, 8), Y = rep(0:1, 8)), iter = 4000, refresh = 0) |>
  query_model("Y[X = 1] > Y[X = 0] :|: X == 1 & Y == 1 & M == 1", 
  using = "posteriors", case_level = c(TRUE, FALSE)) |>
  plot()



```
```{r}
#| label: batch-query
#| echo: true
#| eval: true
#| message: false

models <- list(
  Unrestricted = lipids_model |> 
    update_model(data = lipids_data, refresh = 0),
  Restricted = lipids_model |> set_restrictions("X[Z = 1] < X[Z = 0]") |> 
    update_model(data = lipids_data, refresh = 0)
)


```
```{r}
#| label: batch-query-fig
#| echo: true
#| eval: true

queries <- 
  query_model(
    models, 
     query = list(
       ATE = "Y[X=1] - Y[X=0]", 
       POS = "Y[X=1] > Y[X=0] :|: Y==1 & X==1"),
    case_level = c(FALSE, TRUE), 
    using = c("priors", "posteriors"),
    expand_grid = TRUE)



```
```{r}
#| label: tbl-batch-query
#| tbl-cap: "Results for two queries on two models."
#| echo: true
#| eval: true
#| message: false

queries |>
  dplyr::select(-starts_with("cred"), -sd) |> 
  kabble(align = "c") 



```
```{r}
#| include: true
```

default plot associated with this query:

```{r}
#| label: fig-batch
#| fig-cap: "Default plotting for a a set of queries over multiple models."
#| fig-width: 8 
#| fig-height: 4 
#| echo: true
#| eval: true
#| message: false

queries |> plot()



```
```{r}
#| include: true
```

# Appendix A: Parallelization

```{r}
#| include: true
library("parallel")
options(mc.cores = parallel::detectCores())


```
```{r}
#| echo: true
#| include: true
#| 
library("future")
library("future.apply")

chains <- 3
cores <- 8

future::plan(list(
  future::tweak(future::multisession, 
    workers = floor(cores/(chains + 1))),
  future::tweak(future::multisession, workers = chains)))

model <- make_model("X -> Y")

data <- list(data_1 = data.frame(X = 0:1, Y = 0:1), 
  data_2 = data.frame(X = 0:1, Y = 1:0))

results <- future.apply::future_lapply(data, function(d) {
  update_model(model = model, data = d, chains = chains, refresh = 0)},
 future.seed = TRUE)


```
```{r}
#| include: true
```

# Appendix B: Stan code

```{r}
#| echo: true
#| include: true

CausalQueries:::stanmodels$simplexes



```
```{r}
#| include: true
```

# Appendix C: Benchmarks

```{r}
#| label: tbl-bench1
#| tbl-cap: "Benchmarking 1."
#| echo: true
#| eval: true
#| message: false

iter  <- 4000
times <- 5

model <- list(
  CausalQueries::make_model("X -> Y"),
  CausalQueries::make_model("X1 -> Y <- X2"),
  CausalQueries::make_model("X1 -> Y; X2 -> Y; X3 -> Y")
)

data <- expand_grid(X1 = 0:1, X2 = 0:1, X3 = 0:1) |> uncount(20) |>
  mutate(Y = 1 * (X1 - X2 * X3 + rnorm(n()) > 0))


benchmark_model <- microbenchmark::microbenchmark(
  m1 = CausalQueries::update_model(model[[1]], data, iter = iter, refresh = 0),
  m2 = CausalQueries::update_model(model[[2]], data, iter = iter, refresh = 0),
  m3 = CausalQueries::update_model(model[[3]], data, iter = iter, refresh = 0),
  times = times
)

benchmark_1 <- data.frame(
Model = c(
  "X1 → Y",
  "X1 → Y; X2 → Y",
  "X1 → Y; X2 → Y; X3 → Y"
  ),
  x = c(6, 20, 262),
  t = (benchmark_model |> summary() |> pull(mean))) 

benchmark_1 |> 
 kabble(col.names = c("Model", "Number of parameters", "Runtime (seconds)"), align = "c")  




```
```{r}
#| label: tbl-bench2
#| tbl-cap: "Benchmarking 2."
#| echo: true
#| message: false
#| eval: true

model <- CausalQueries::make_model("X -> Y")

data <- lapply(10^c(1:4), function(n) {
  CausalQueries::make_data(model, n)
})

benchmark_data <- microbenchmark::microbenchmark(
  d0 = CausalQueries::update_model(model, data[[1]], iter = iter, refresh = 0),
  d1 = CausalQueries::update_model(model, data[[2]], iter = iter, refresh = 0),
  d2 = CausalQueries::update_model(model, data[[3]], iter = iter, refresh = 0),
  d3 = CausalQueries::update_model(model, data[[4]], iter = iter, refresh = 0),
  times = times
)

benchmark2 <- data.frame(
  Model = c("X1 → Y", 
            "X1 → Y", 
            "X1 → Y", 
            "X1 → Y"),
  x = c(10, 100,  1000, 10000),
  t = (benchmark_data |> summary() |> pull(mean))
) 

benchmark2 |>
 kabble(col.names = c("Model", "Number of observations", "Runtime (seconds)"), align = "c")
```

